A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-07 doc/popl2013/langdef.tex
more updates

Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
include booleans, integers, tuples, and pointers, which are annotated with a list
of regions into which they may point.  There is a special type for
{\em colorings}, which are used to specify partitions of regions into
subregions.  Functions/tasks in Legion are named, accept one or more
arguments and return a value of some type.  A function also declares
the region access privileges and coherence it requires.  Function
types are universally quantified over all logical region bindings,
with all argument and result types, as well as the privileges and
coherence requirements being expressed relative to the quantified
region names.  The final additional value type that Legion uses is a
region relation, which stores a value and one or more logical regions
satisfying associated region constraints.  An instance of a region
relation can be read out of the heap and unpacked, giving new local
names to the logical regions contained in the instance.
% and tracking
%the constraints between those new logical regions.

@@

Core Legion has no subtyping, but a pointer's region can be changed via explicit
{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
but {\em downregion} must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge id \in \mbox{\em functions}(\Gamma) \Rightarrow reduces_{id}(r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

where: 
\begin{itemize}
\item $M$ is the mapping from logical regions $r_i$ in types to physical regions $\rho_i$, which are sets of memory locations.

It will be useful to apply this logical-to-physical mapping to other
things that defined in terms of logical regions to yield a corresponding definition in terms of
physical regions.  The notation $\hat x = M \llbracket x \rrbracket$ is used to describe the
result of replacing all free logical region names $r_i$ in $x$ with the corresponding physical
region $\rho_i$.  (Note that $M \llbracket x \rrbracket$ is not
well-formed if it contains a free logical region name $r_j \not\in dom(M)$.)
\item $L$ is the mapping of local variable names to their values on the stack.  These values are
immutable, and private to the function - it is not possible to get a pointer to a local value.
\item $H$ is the heap typing, mapping from memory locations to types.  The heap typing does not
change during the execution of a legion program.  The Legion type system is strong enough to
support the use of garbage collection
to allow the reuse of a memory location (potentially with a different type stored in that
location) once all existing references to it are dead, but we
do not consider that further in this paper.
\item $S$ is the store, mapping from memory locations $l_i$ to the current value stored.
\item $C$ is the ``clobber set'', a set of memory locations that may be accessed by concurrently executing operations (??? does this need to be split into read/write/reduce?)
\item $e$ is the expression being evaluated
\item $v$ is the result of the evaluation
\item $E$ is a memory trace (i.e. the effects) of the execution, a list of memory operations
in the order they were performed.  An operation $\epsilon$ is one of the following:
\begin{itemize}
\item $read(l, c, v)$, a read of location $l$ with coherence mode $c$, returning value $v$
\item $write(l, c, v)$, a write to location $l$ with coherence mode $c$, storing value $v$
\item $reduce_{id}(l, c, v)$, a reduction into location $l$ with coherence mode $c$ using a
commutative function $id$ with value $v$
\end{itemize}
where coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).

The following helper functions are used in some of the rules that deal with memory effects:
\begin{itemize}
\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
constrain how the operations in that trace can be interleaved with other sibling tasks' operations
\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
interleave the memory traces of two or more subexpressions
\end{itemize}
Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
the {\em valid\_interleave} predicate, we leave the precise definition to a later section.  For this
section, we need only one of the necessary conditions of a valid interleaving, that a valid
interleaving is at least an interleaving - it contains all the memory operations from each of the
constituent traces and no more, and that each trace's operations remain in order in the interleaved
trace.
\end{itemize}

There are a number of rules that appear to be nondeterministic, allowing multiple choices for
valid evaluation.  In reality, only the memory allocation case (i.e. new $T@r$) has a truly
nondeterministic choice (i.e. which currently-unused memory location of the right type to 
use).  A particular memory allocation algorithm is likely to be deterministic, but the
design of such algorithms is unrelated to the subject at hand, and by abstracting the location 
choice, we show that our results hold for any correct memory allocation algorithm.  (In fact,
our criteria are even weaker than that.  Our soundness result even holds if the location returned
by ``new $T@r$'' is one that is already in use, as long as the heap typing of that location is
correct.)

The remaining cases of apparent nondeterminism are the result of how we handle the semantics of
concurrent execution.  An operational semantics that explicitly tracks the state of all 
concurrently executing operations would allow precise specification of the result of each 
evaluation step, but would be extremely cumbersome.  Instead, we identify the cases in which
nonlocal effects (i.e. those that result from the concurrent evaluation of other expressions)
can impact the local evaluation, and make a ``guess'' of what value would result from an 
operation.  The check that the guess was consistent with the state of the system is deferred until
the previously nonlocal effects become local (i.e. contained within the evaluation of an enclosing
expression tree).  If the guess is not consistent, then the attempted proof of the subexpression's
behavior is flawed and must be reattempted.

In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
evaluation to guess any value as long as it is of the right type.  The consistency of the value
is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
freedom of choice, deferring some of the consistency checks to the corresponding check further up
the expression tree.  The final case of apparent nondeterminism is in the choice of what initial
locations to include in the result of a ``newcolor $r$'' expression.  The inclusion of locations
beyond those which are explicitly requested by the application is necessary for the success of
subsequent memory allocation within a subregion created from a partitioning using the coloring, but
that allocation might not occur within the {\em partition} expression's
body.  Although the correctness of these partitioning guesses cannot be fully verified until the
entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
it the right type in $H$) rather than having to start the proof over from the beginning.

Most of the heavy lifting in the Legion core language occurs in the function call expression.
The expressions for all the function's arguments are evaluated in the caller's context, but a
new context is assembled for the evaluation of the function's body.  The region mapping used
in the callee only uses the caller's mapping indirectly - the only logical region names available
are those over which the called function is quantified.  Similarly, the set of local values is
replaced with a mapping of formal parameter names to the values of the arguments.  The store
is updated to account for side-effects of those argument evaluations.  Finally, if the function
being called operates with relaxed coherence (i.e. either {\em atomic} or {\em simult} instead of
{\em excl}), the locations with relaxed coherence are added to the clobber set, allowing the
impact of concurrent modifications of the store to be handled by the caller.  This is done by
annotating the called function's memory trace with the relaxed coherence on a per-location basis
and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.

@@

\opsenvx e_1 \mapsto l, E \\

@@

{\opsenvx \text{read}(e_1) \mapsto v, E \dplus [ read(l, excl, v) ]}

@@

An important property of the legion type system is that any expression that is well-typed is
guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
A judgement $M \vdash E : \Phi$ holds if all the memory operations in $E$ are of the types and
to locations covered by the privileges in $\Phi$:

\begin{center}
\begin{tabular}{l@{ }l@{ }l}
$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
& & $(\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)$
\end{tabular}
\end{center}

The proof that $M \vdash E : \Phi$ holds for all well-typed Legion expressions is achieved by
induction over the structure of the expression.  Soundness is contingent on the following
consistent properties, which must be shown to hold for each subexpression:
\begin{itemize}
\item $M \sim \Omega$, a region mapping must satisfy the region constraints under which the expression was typed
\item $L \sim_H \Gamma$, the local values must have types consistent with the environment under which the expression was typed
\item $S \sim H$, the initial state of the store must contain values whose types are consisten with the heap typing
\end{itemize}

The proof requires two additional properties be proved for each subexpression:
\begin{itemize}
\item $v \sim_H M \llbracket T \rrbracket$, a subexpresion must evaluate to a value of the right type
\item $E \sim H$, the values written to the store must match the heap types of the location to which they're stored
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these properties.

@@

\begin{thm}
Let $e$ be a well-typed Legion expression.  Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $v$ and $E$ are the value and memory trace resulting from a permissable terminating evaluation of
$e$, then:
$$
\begin{array}{l}
v \sim_H M \llbracket T \rrbracket \\
E \sim H \\
M \vdash E : \Phi
\end{array}
$$
\end{thm}

Due to space constraints, we limit the discussion to the general strategy of the proof.  The full
proof can be found here\cite{LegionTypes12}.  If $v$ and $E$ are the result of a permissible evaluation of
$e$, then a proof exists of:
$$\opsenv{e}{v}{E}$$
The soundness of the result is shown by induction over the structure of this proof.  For each of 
the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
evaluations (provided we can show the consistency of the mappings and store used by that
subexpression) and prove that the result of the expression's evaluation is sound.  Although there
are a multitude of cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The proofs of these lemmas can also be found here\cite{LegionTypes12}.  The proof of the theorem is organized
into six pieces, one for each of the three consistency properties needed for subexpressions and one
for each of the three soundness properties we wish to show.

@@

\item Only three expressions have subexpressions that modify $M$ or $\Omega$ - those that don't
trivially satisfy the region mapping consistency requirement.  For the
{\em partition} expression, the assumed consistency of the coloring can be shown to preserve
the region mapping consistency with respect to the constraints.  Similarly, the consistency of 
a region relation instance guarantees the maintenance of consistency in an {\em unpack}
expression.  Finally, the body of a called function executes with a different region mapping,
but with an initially-empty set of constraints, so they are trivially satisfied.

\item Only three expression have subexpressions that modify $L$, $\Gamma$, $M$.  The
{\em partition} expression only modifies $M$, and the requirement that it not reuse existing
names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  The value and type of the
binding created in a {\em let} expression is obviously consistent, while the binding created in
an {\em unpack} expression is less obviously so, requiring an induction over the type of the 
unpacked value to show equivalence under the new mapping.  The last case is again the body of a
called function, which requires the same style of proof as for the {\em unpack} expression to
be performed for each formal parameter.

\item The heap typing consistency of all stores used in subexpressions follows directly from
Lemma~\ref{lemma:heapconst:apply}.

\item With the necessary consistency conditions all satified, the inductive hypothesis asserts
the type consistency of all subexpression result values.  The consistency of the result of 
the {\em upregion} is guaranteed by the type checking requirement of appropriate subregion
constraints and the mapping's consistency with those constraints, whereas {\em downregion}
expression's result is consistent due to the run time check that is performed.  (The requirement
for appropriate constraints in the {\em downregion} typing rule is primarily for performance
(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
work at runtime.)

The consistency of a {\em read} expression's result is trivial when an address falls in the
clobber set, and counts on the consistency of the store otherwise.  The consistency of a 
{\em color} expression's result depends on the pointer subexpression's consistency and the 
removal of any previous coloring of that location from the coloring set.  The remaining
interesting cases arise from changes to the mapping $M$ rather than transformations on the
value $v$.  In the case of {\em partition} and {\em unpack} expressions, the type system's
requirement that the body subexpression's result not use the regions that were added to the
mapping allow the changes to the mapping to be ignored.  The last case is, yet again, the body
of a called function, and the same strategy that was used for the type consistency of the
formal parameters works in reverse for the function's result.

\item The type consistency of the value in an expression's effect follows from
Lemmas~\ref{lemma:heapconst:effects1} and~\ref{lemma:heapconst:effects2}.  New effects are
added by the {\em write} and {\em reduce} expressions, but their values' consistency follows
directly from the inductive hypothesis on their subexpressions.  Finally, the consistency of the
values in a called function's memory trace is addressed in the same way as the return value.

\item The proof of the crucial property of containment of effects within the available privileges
is very similar to the previous step, with Lemmas~\ref{lemma:effsound:effects1}
and~\ref{lemma:effsound:effects2} covering most cases, straightforward proofs for the {\em read},
{\em write}, and {\em reduce} expressions, and one final special case for function calls.  The 
inductive hypothesis asserts that the function body's effects are contained within its own
privileges.  Those are a subset of the caller's $\Phi^*$, and then Lemma~\ref{lemma:phiclosure}
is used to complete the proof.
\end{itemize}

@@

\subsection{Coherence}

The soundness proof above guarantees that the evaluation of an expression produces a value of the
right type, and do so within the regions of memory for which privileges are held, but for the
Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
produces the (or at least, a) correct value.  The key to doing this is to place additional 
constraints on what it means to be a valid interleaving of two or more memory traces.  In
addition to a valid interleaving being an interleaving of the individual traces, we will 
require that it be:
\begin{itemize}
\item {\em coherent} - each memory read operation should see the effect of all writes and reductions 
that occur earlier in the interleaved trace to the same location, and
\item {\em sequentially equivalent} - the value read by each memory read operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid {\em program order}.
\end{itemize}
These requirements are relaxed in some cases due to either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or due to coherence
modes other than {\em excl} on operations with the traces (representing operations that may 
occur concurrently as part of evaluating other parts of the same function).

@@

The second case is a subset of the first, in which all apparently-interfering operations have
the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
We again modify the noninterference criteria for effects, traces, and privileges, this time
only allowing $simult$ coherence annotations, and can show that subexpression with privileges
that can be shown to be ``simulataneously noninterfering'' (under mapping $M$) can be executed
in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
execution.  Again, a special (but important) case exists where all privileges for each
subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
parallel execution to be determined at compile time.

>>>>

Figure~\ref{fig:langdef} defines the core subset of Legion.  Types
include booleans, integers, tuples, and pointers.  Note that pointers
are annotated with a list of regions into which they may point; in a
well-typed program, the pointer is guaranteed to point to at least one
of those regions. There is a special type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions/tasks in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  Function types are
universally quantified over all region names appearing in the type.
The final value is a region relation, which stores a value and one or
more logical regions satisfying associated region constraints.  An
instance of a region relation can be read out of the heap and
unpacked, giving new local names to the logical regions contained in
the instance.
% and tracking
%the constraints between those new logical regions.

@@

Core Legion has no subtyping. A pointer's region is changed only via explicit
{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
but {\em downregion} must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \Rightarrow reduces_{id}(r) \in \Phi^* \\
\ \ \ \mbox{for every function identifier $id$}
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

where: 
\begin{itemize}
\item $M$ is a mapping from logical regions $r_i$ to {\em physical regions} $\rho_i$, which are sets of concrete memory locations.
We write $\hat x = M \llbracket x \rrbracket$ for the application of $M$ to other objects (e.g., types), 
replacing all free logical region names $r_i$ in $x$ with corresponding physical regions $\rho_i$.  
(Note $M \llbracket x \rrbracket$ is ill-formed if it contains a free logical region name $r_j \not\in dom(M)$.)

\item $L$ is a mapping of local variables to (immutable) values.

\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.

\item $S$ is a store, mapping memory locations $l_i$ to values.

\item $C$ is a {\em clobber set}, memory locations that may be accessed by concurrently executing operations.

\item $e$ is an expression.

\item $v$ is the result of evaluatinig $e$.

\item $E$ is a {\em memory trace} of the execution, a list of {\em memory operations}
in the order performed.  A memory operation $\epsilon$ is either
$read(l, c, v)$ (a read of location $l$ with coherence mode $c$, returning value $v$),
$write(l, c, v)$ (a write to location $l$ with coherence mode $c$, storing value $v$),
or $reduce_{id}(l, c, v)$ (a reduction into location $l$ with coherence mode $c$ using an
associative function $id$ with value $v$).  Coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
\end{itemize}

Helper functions are used in some rules:
\begin{itemize}
\item $apply(S, E) = S'$ computes the store $S'$ resulting from applying memory trace $E$ to initial store $S$.

\item $mark\_coherence(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $Q$.

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subequence and no other operations.

Some rules modeling concurrent execution are locally
non-deterministic, guessing values that result from
concurrent tasks.  For example, the rule for {\tt read(e)} returns an
arbitrary value $v'$ of the correct type if the location being read is
in the clobber set $C$, the set of locations concurrent tasks may
clobber (i.e., update).\footnote{Note that this is also the reason for
including a global heap typing.}  The guess is eventually checked for
consistency by the $valid\_interleave$ predicate of some enclosing
expression, at a point where the task that wrote $v'$ is also local to
the expression being evaluated.

Another prescient operation is {\tt newcolor r}, which guesses initial
memory locations and their colors for a partition.  These locations
support partition-then-allocate, ensuring the memory mappping $M$ has
locations for the new subregions that can be allocated by {\tt new}.
As is usual, memory allocation is also non-deterministic, choosing any
location of the correct type in the region.  For simplicity we do not
require locations to be ``fresh''; our soundness results hold even if
{\tt new T@r} returns locations already in use, provided the heap
typing of the location is correct.

The most involved rule in Core Legion's semantics is for task calls.
Note that the only regions accessible to the callee (through $M'$) are the region arguments.
Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
{\em excl}), those locations are added to the clobber set, allowing the
impact of concurrent modifications of the store to be handled by the caller.  
%This is done by
%annotating the called function's memory trace with the relaxed coherence on a per-location basis
%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.

@@

@@

{\opsenvx \text{read}(e) \mapsto v, E \dplus [ read(l, excl, v) ]}

@@

A key property of the Legion type system is that any expression that is well-typed is
guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
A judgement $M \vdash E : \Phi$ holds if all the memory operations in memory trace $E$ are of the types and
to locations covered by the privileges in $\Phi$:
\[ 
\begin{array}{rl}
M \vdash & E : \Phi \Leftrightarrow \forall \epsilon \in E. \\
& (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
& (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
& (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
\end{array}
\]

Our goal is to show $M \vdash E : \Phi$ holds for the execution of an expression $e$ with static permissions $\Phi$.
As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
For our results, three consistency properties are needed:
\begin{itemize}
\item $M \sim \Omega$, a region mapping satisfies the region constraints under which the expression was typed
\item $L \sim_H \Gamma$, local values have types consistent with the environment under which the expression was typed
\item $S \sim H$, the initial state of the store contains values whose types are consisten with the heap typing
\end{itemize}

Two additional properties are proven for each subexpression:
\begin{itemize}
\item $v \sim_H M \llbracket T \rrbracket$, the expression yields a value of the right type
\item $E \sim H$, values written to the store match the heap types of their locations
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these properties.

@@

\begin{thm}
\label{thm:effects}
\rm
If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $M \sim \Omega$, $L \sim_H \Gamma$ and $S \sim H$,
then
$v \sim_H M \llbracket T \rrbracket$, $E \sim H$ and $M \vdash E : \Phi$.
\end{thm}

We spare the reader the lengthy proof, which can be found in
\cite{LegionTypes12}, and merely outline the general strategy.  The
proof is a standard induction on the structure of the derivation in
the operational semantics.  The proof is organized into six pieces,
one for each of the three consistency properties that must be shown
for the execution environment of subexpressions and one for each of
the three properties of the execution's results we wish to show.

\texcomment{
For each of 
the 23 different forms a Legion expression can have, we assume the soundness of the subexpressions'
evaluations (provided we can show the consistency of the mappings and store used by that
subexpression) and prove that the result of the expression's evaluation is sound.  Although there
are a multitude of cases to be proven, many of them are similar, and benefit from
the use of the following lemmas:

@@

The proofs of these lemmas can also be found here\cite{LegionTypes12}.  
}

@@

\item[$M \sim \Omega$] Three expressions have subexpressions that modify $M$ or $\Omega$ and therefore do not
trivially satisfy region mapping consistency.  For {\tt partition},
the consistency of the coloring preserves region mapping consistency
with respect to the constraints.  For {\tt unpack}, the consistency of
a region relation instance guarantees consistency of region mapping.
Finally, the body of a called function with an initially-empty set of
constraints, which are trivially satisfied.

\item[$L \sim_H \Gamma$] Four expressions have subexpressions that modify $L$, $\Gamma$, or $M$.  For
{\tt partition}, which only modifies $M$, the requirement that it not reuse existing
names ensures that $M \llbracket \Gamma \rrbracket$ does not change.  For {\tt let},
the value and type of the binding is obviously consistent, while the binding created in
an {\tt unpack} is less obviously so, requiring an induction over the type of the 
unpacked value to show equivalence under the new mapping.  The last case is the body of a
called function, which requires the same style of proof as for {\tt unpack} 
for each formal parameter.

\item[$S \sim H$] The heap typing consistency of all stores used in subexpressions follows directly from the following
lemma: If $S \sim H$ and $E \sim H$, then $\text{apply}(S, E) \sim H$.

\end{itemize}

With the necessary consistency conditions proven for a subexpression, the conclusion of the theorem follows for
the subexpression by the inductive hypothesis.  The remaining obligation is to prove the the three parts of the
conclusion for an entire expression:
\begin{itemize}

\item[$v \sim_H M \llbracket T \rrbracket$]    The consistency of 
{\tt upregion} is guaranteed by the type checking requirement of appropriate subregion
constraints and the mapping's consistency with those constraints, and {\tt downregion}'s
result is consistent because of the run time check.  
%(The requirement
%for appropriate constraints in the {\em downregion} typing rule is primarily for performance
%(i.e. knowing which part of the region tree to examine) and to catch downcasts that will never
%work at runtime.) The consistency of a {\tt read}'s result is trivial for an address in the
clobber set and uses the consistency of the store otherwise.  The consistency of a 
{\tt color}'s result depends on the pointer subexpression's consistency and the 
removal of any previous coloring of that location from the coloring set.

The remaining interesting cases arise from changes to the mapping $M$ rather than transformations on the
value $v$.  In the case of {\tt partition} and {\tt unpack}, the type system
guarantees that the subexpression's result cannot use the regions that were added to the
mapping, allowing the changes to the mapping to be ignored.  The last case is, yet again, the body
of a called function, and the same strategy that was used for the type consistency of the
formal parameters works in reverse for the function's result.

\item[$E \sim H$] The type consistency of the value in an expression's effect follows from two lemmas applied to the effects
of subexpressions:
if $E_1 \sim H$ and $E_2 \sim H$, then $E_1 \dplus E_2 \sim H$; and 
if $E_1 \sim H$ and $E_2 \sim H$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $E' \sim H$.
New effects are added by {\tt write} and {\tt reduce} expressions, but consistency follows
directly from the induction hypothesis.  Finally, the consistency of the
values in a called function's memory trace is addressed in the same way as the return value.

\item[$M \vdash E : \Phi$] The proof of the crucial property of containment of effects within the available privileges
is similar in outline to the previous step.  The easy cases are covered by lemmas combining the corresponding results
for subexpressions: if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$, then $M \vdash E_1 \dplus E_2 : \Phi$, and
if $M \vdash E_1 : \Phi$ and $M \vdash E_2 : \Phi$ and $valid\_interleave(S, C, E', E_1, E_2)$, then $M \vdash E' : \Phi$.
Straightforward proofs cover {\tt read}, {\tt write}, {\tt reduce}, and one final special case for function calls. 
\end{itemize}

@@

\subsection{Coherence}
\label{sec:coherence}

Theorem~\ref{thm:effects} guarantees expression evaluation produces a value of the
correct type and does so within the regions of memory for which privileges are held.

 but for the
Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
produces the (or at least, a) correct value.  The key to doing this is to place additional 
constraints on what it means to be a valid interleaving of two or more memory traces.  In
addition to a valid interleaving being an interleaving of the individual traces, we will 
require that it be:
\begin{itemize}
\item {\em coherent} - each memory read operation should see the effect of all writes and reductions 
that occur earlier in the interleaved trace to the same location, and
\item {\em sequentially equivalent} - the value read by each memory read operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid {\em program order}.
\end{itemize}
These requirements are relaxed in some cases due to either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or due to coherence
modes other than {\em excl} on operations with the traces (representing operations that may 
occur concurrently as part of evaluating other parts of the same function).

@@

The second case is a subset of the first, in which all apparently-interfering operations have
the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
We again modify the noninterference criteria for effects, traces, and privileges, this time
only allowing $simult$ coherence annotations, and can show that subexpression with privileges
that can be shown to be ``simultaneously noninterfering'' (under mapping $M$) can be executed
in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
execution.  Again, a special (but important) case exists where all privileges for each
subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
parallel execution to be determined at compile time.
